---
title: "**Assignment 7: Ensemble Tree-based Methods**"
output: html_document
---

#### **Topic: Predicting 30-Day Hospital Readmission After Myocardial Infarction Using Ensemble Methods**

In class today, we worked on a Group Exercise to sketch out an analytic pipeline to predict
readmission for myocardial infarction. In the assignment, you will implement your pipeline
(or a simplified version of it) using a sample dataset.

The dataset (`mi.data`) consists of an ID variable, 14 features describing clinical tests
and comorbidities, and an indicator for whether the individual was readmitted to the hospital
within 30 days of discharge.

**Feature Information:**

* `readmission`:   Readmitted within 30 days — 1: Yes, 0: No (Outcome) [Categorical]
* `Age`:           Age at initial MI in years [Continuous]
* `Sex`:           Reported by patient — 0: Male, 1: Female, 2: Non-binary/Other [Categorical]
* `sodium`:        Serum sodium (mmol/L) [Continuous]
* `ALT`:           Liver enzymes (IU/L) [Continuous]
* `WBC`:           White blood cell count (billions/L) [Continuous]
* `ESR`:           Erythrocyte sedimentation rate [Continuous]
* `SBP`:           Systolic blood pressure at intake (mmHg) [Continuous]
* `DBP`:           Diastolic blood pressure at intake (mmHg) [Continuous]
* `Pulm.adema`:    Pulmonary edema (1=Yes, 0=No) [Categorical]
* `FC`:            Functional class of angina pectoris in the last year (0–4) [Categorical]
* `Arr`:           Presence of arrhythmia (1=Yes, 0=No) [Categorical]
* `Diab`:          Presence of diabetes (1=Yes, 0=No) [Categorical]
* `Obesity`:       Presence of obesity (1=Yes, 0=No) [Categorical]
* `Asthma`:        Presence of asthma (1=Yes, 0=No) [Categorical]

---

#### **0. Data Preparation**

**Set up needed libraries**

```{r setup_library, message=FALSE, warning=FALSE}
library(tidyverse)
library(caret)
library(randomForest)
library(gbm)
library(pROC)

set.seed(123)
```

**Import dataset**

```{r import, message=FALSE}
mi_df = read_csv(file = "./mi.data.csv")

mi_df = mi_df |> select(-ID)

mi_df
```

**Clean dataset**

```{r clean, message=FALSE}
# Check for missing values across all variables before proceeding
miss_count = colSums(is.na(mi_df))
cat("Missing values per variable:\n")
print(miss_count)

# Retain only complete cases
mi_df = mi_df[complete.cases(mi_df), ]
cat("\nTotal observations after removing missing:", nrow(mi_df), "\n")

# Convert binary and ordinal clinical features to factors.
# These are coded as integers (0/1) but represent categories, not quantities.
# Leaving them numeric would imply a false ordering that could mislead the models.
mi_df = mi_df |>
  mutate(
    Sex        = as.factor(Sex),
    Pulm.adema = factor(Pulm.adema, levels = c("0", "1")),
    FC         = factor(FC,         levels = c("0", "1", "2", "3", "4")),
    Arr        = factor(Arr,        levels = c("0", "1")),
    Diab       = factor(Diab,       levels = c("0", "1")),
    Obesity    = factor(Obesity,    levels = c("0", "1")),
    Asthma     = factor(Asthma,     levels = c("0", "1")),
    # Outcome: "No" set as reference level; "Yes" is the positive (event) class
    readmission = factor(ifelse(readmission == 1, "Yes", "No"),
                         levels = c("No", "Yes"))
  )

mi_df
```

**Overview of continuous features**

```{r descriptives, message=FALSE}
mi_df |>
  summarise(across(where(is.numeric),
                   list(Mean = ~round(mean(.), 2),
                        SD   = ~round(sd(.), 2),
                        Min  = ~round(min(.), 2),
                        Max  = ~round(max(.), 2)))) |>
  pivot_longer(everything(),
               names_to  = c("Variable", "Stat"),
               names_sep = "_(?=[^_]+$)") |>
  pivot_wider(names_from  = Stat,
              values_from = value) |>
  knitr::kable(caption = "Table 1: Descriptive Statistics of Continuous Features")
```

**Check outcome balance**

```{r balance, message=FALSE}
balance_tbl = mi_df |>
  count(readmission) |>
  mutate(Percent = round(100 * n / sum(n), 1))

balance_tbl |>
  knitr::kable(caption = "Table 2: Frequency and Proportion of Readmission Outcome")
```

**NOTE:** From **Table 2**, we observe a substantial class imbalance in the outcome
variable, with 90.6% of patients (n = 1,541) classified as not readmitted and only 9.4%
(n = 159) readmitted within 30 days of discharge. To prevent models from being biased
toward always predicting the majority class (no readmission), **down-sampling** will be
applied within each cross-validation fold. This corrects for the imbalance during training
without leaking information from held-out folds into the resampling process, and ensures
that sensitivity — the model's ability to correctly identify patients who will truly be
readmitted — is not sacrificed in favor of inflated overall accuracy.

**Partition the data into training and testing using a 70/30 split**

```{r split, message=FALSE}
set.seed(123)

train_indices = createDataPartition(y = mi_df$readmission, p = 0.7, list = FALSE)
train_data    = mi_df[ train_indices, ]
test_data     = mi_df[-train_indices, ]

split_counts = tibble(
  Dataset = c("Training data", "Testing data"),
  N       = c(nrow(train_data), nrow(test_data)),
  Percent = round(100 * N / nrow(mi_df), 1)
)

split_counts |>
  knitr::kable(caption = "Table 3: Number of Observations in Training and Test Sets")
```

**NOTE:** A stratified 70/30 split using `createDataPartition()` preserves the proportion
of readmitted patients in both sets, yielding 1,191 observations in the training set and
509 in the test set. The test set is set aside completely and will not be examined until
final model evaluation in Section 2. All model selection decisions are made using
cross-validation performance on the training set only.

All 14 features are retained for modeling based on the following reasoning:

* `Age`, `SBP`, `DBP`, `sodium`, and `WBC` are established clinical indicators of acute
  cardiac condition severity and physiological stress at admission.
* `ALT` and `ESR` capture liver function and systemic inflammation, both associated with
  post-AMI recovery trajectory and complication risk.
* `Pulm.adema`, `FC`, and `Arr` reflect direct cardiac complications at admission that
  are known predictors of adverse post-discharge outcomes.
* `Diab`, `Obesity`, and `Asthma` are comorbidities that increase readmission risk by
  complicating recovery and elevating vulnerability to secondary events.
* `Sex` is retained to capture potential differences in clinical presentation and outcomes
  across patient groups.

---

#### **1. The Comparison of Two Models**

Construct **two prediction models** to predict 30-day readmission using all features. Each
model is optimized via cross-validation on the *training data*, and performance is compared
before any evaluation on the test set. The two algorithms used are:

* **a) Random Forest** — an ensemble of decision trees built on bootstrapped samples, where
  each tree uses a random subset of features at each split to reduce correlation between trees.
* **b) Gradient Boosting Machine (GBM)** — an ensemble that grows trees *sequentially*, where
  each new tree corrects the errors (pseudo-residuals) of the prior model, learning slowly
  and deliberately to improve prediction.

First, a shared 10-fold cross-validation control is created for use across both models.
*Down-sampling* is applied within each fold to correct for class imbalance in the outcome.

```{r control, message=FALSE}
train_control = trainControl(
  method          = "cv",
  number          = 10,
  sampling        = "down",        # down-sample majority class within each fold
  classProbs      = TRUE,          # needed to generate predicted probabilities for ROC
  savePredictions = "final"        # saves fold-level predictions for confusionMatrix()
)
```

---

**a) Random Forest**

A random forest model that tunes `mtry` — the number of features eligible for selection
at each split — via 10-fold cross-validation.

```{r rf, message=FALSE}
set.seed(123)

p = ncol(train_data) - 1   # total number of predictors (all columns minus outcome)

# Three mtry values tested, following the lecture demo approach:
#   p       → equivalent to bagging (all features at each split, trees are correlated)
#   sqrt(p) → standard recommendation for RF classification tasks
#   0.5*p   → intermediate de-correlation between bagging and full RF
mtry_vals = c(p, round(sqrt(p)), round(0.5 * p))
mtry_grid  = expand.grid(.mtry = mtry_vals)

model_rf = train(
  readmission ~ .,
  data      = train_data,
  method    = "rf",
  metric    = "Accuracy",
  tuneGrid  = mtry_grid,
  trControl = train_control,
  ntree     = 500            # 500 trees — RF does not overfit with more trees
)

# Best mtry value selected by cross-validation
model_rf$bestTune

# CV accuracy for each mtry value tested
model_rf$results

# Variable importance: accuracy-based OOB importance
# Measures how much prediction accuracy drops when each variable is randomly permuted
varImp(model_rf)

plot(varImp(model_rf), top = 10,
     main = "Figure 1: Random Forest — Top 10 Features by Variable Importance")

# Both MeanDecreaseAccuracy and MeanDecreaseGini displayed side by side
varImpPlot(model_rf$finalModel,
           main = "Figure 2: Random Forest — Variable Importance (Both Measures)")

# Cross-validated confusion matrix from training folds
confusionMatrix(model_rf)
```

**Description:** The optimal `mtry` value selected via cross-validation was 4 —
approximately √p where p = 14 predictors — yielding a mean CV accuracy of 57.25% and
Kappa of 0.062 after down-sampling. As shown in the results table, `mtry` = 4
outperformed both `mtry` = 7 (accuracy = 56.01%) and `mtry` = 14 (accuracy = 55.74%),
suggesting that restricting features at each split effectively de-correlates trees and
reduces variance compared to bagging, consistent with the standard recommendation for
classification tasks. As shown in _Figure 1_ and _Figure 2_, `WBC` (importance = 100.0)
and `Age` (importance = 86.5) ranked as the two most important predictors in both
importance measures, followed by `sodium` (76.8) and `ESR` (73.0), suggesting that acute
physiological stress markers at admission and patient age are the strongest drivers of
readmission risk. The cross-validated confusion matrix shows the model correctly identified
5.6% of true readmission cases and 51.6% of non-readmission cases on average across folds.

---

**b) Gradient Boosting Machine (GBM)**

A GBM model that tunes four hyperparameters via 10-fold cross-validation: `n.trees`
(number of trees), `shrinkage` (learning rate λ), `interaction.depth` (tree depth d),
and `n.minobsinnode` (minimum observations per terminal node).

```{r gbm, message=FALSE, warning=FALSE}
set.seed(123)

# Hyperparameter tuning grid:
#   n.trees:           number of sequential boosting iterations
#   shrinkage:         learning rate — smaller = slower learning, requires more trees
#                      0.01 and 0.001 are typical starting values from lecture
#   interaction.depth: d=1 → stumps (additive model, no interactions) — weak learner approach
#                      d=3 → allows 3-way interactions, more complex
#   n.minobsinnode:    minimum cases per terminal node, controls tree size
grid_gbm = expand.grid(
  n.trees           = c(100, 200, 300, 500),
  shrinkage         = c(0.01, 0.001),
  interaction.depth = c(1, 3),
  n.minobsinnode    = 10
)

model_gbm = train(
  readmission ~ .,
  data         = train_data,
  method       = "gbm",
  distribution = "bernoulli",   # bernoulli distribution for binary classification
  metric       = "Accuracy",
  tuneGrid     = grid_gbm,
  trControl    = train_control,
  verbose      = FALSE          # suppress per-iteration output
)

# Best hyperparameter combination selected by CV
model_gbm$bestTune

# Top CV results — look for selected shrinkage and interaction depth
model_gbm$results |>
  arrange(desc(Accuracy)) |>
  select(n.trees, shrinkage, interaction.depth, Accuracy, Kappa) |>
  head(10)

# Variable importance via gbm summary — relative influence per feature
gbm_imp = summary(model_gbm$finalModel,
                  n.trees = model_gbm$bestTune$n.trees,
                  plotit  = FALSE)

gbm_imp |>
  slice_max(rel.inf, n = 10) |>
  ggplot(aes(x = rel.inf, y = reorder(var, rel.inf))) +
  geom_col(fill = "steelblue") +
  labs(x     = "Relative Influence",
       y     = "Feature",
       title = "Figure 3: GBM — Top 10 Features by Variable Importance") +
  theme_minimal()

# Cross-validated confusion matrix from training folds
confusionMatrix(model_gbm)
```

**Description:** The optimal hyperparameters selected via cross-validation were
`n.trees` = 500, `shrinkage` = 0.01, and `interaction.depth` = 3, yielding a mean CV
accuracy of 55.76% and Kappa of 0.019 after down-sampling. A shrinkage of 0.01 was
selected over 0.001, indicating that a faster learning rate with 500 trees was sufficient
for this dataset — a smaller shrinkage of 0.001 would require substantially more trees to
reach the same level of learning. An interaction depth of 3 was selected rather than 1
(stumps), suggesting the model benefited from capturing interactions among features rather
than purely additive effects. As shown in _Figure 3_, `Age` and `WBC` were the most
influential predictors by relative influence, consistent with the Random Forest results.
The cross-validated confusion matrix shows the model correctly identified 4.6% of true
readmission cases and 51.1% of non-readmission cases on average across folds.

---

#### **2. Model Selection**

Select an "optimal" model and calculate final evaluation metrics in the test set.
**Only one model should be applied to the test set.** What do you conclude about your
final model's performance?

**Compare Models on Training Performance**

```{r comparison, message=FALSE}
results = resamples(list(
  RandomForest = model_rf,
  GBM          = model_gbm
))

summary(results)

bwplot(results,
       main = "Figure 4: 10-Fold CV Performance — Random Forest vs. GBM")
```

**Description:** As shown in the accuracy and Kappa results above and _Figure 4_, Random
Forest achieved the highest mean CV accuracy (57.25%, Kappa = 0.062), compared to GBM
(55.76%, Kappa = 0.019). **Random Forest is selected as the final model** because it
achieved consistently higher cross-validated accuracy and Kappa across all 10 folds,
indicating more stable and reliable discriminative ability. Additionally, Random Forest's
parallel ensemble structure — where trees are grown independently on bootstrapped samples —
provides greater stability with this relatively small clinical dataset (n = 1,700) compared
to GBM's sequential boosting approach, which is more sensitive to hyperparameter choices
and class imbalance.

**Final Model Evaluation (Test Set)**

The final Random Forest model is applied once to the held-out test set to report
unbiased evaluation metrics.

```{r evaluation, message=FALSE}
set.seed(123)

final_model = model_rf

final_pred = predict(final_model, newdata = test_data)

confusionMatrix(final_pred, test_data$readmission, positive = "Yes")
```

```{r roc, message=FALSE}
final_pred_prob = predict(final_model, newdata = test_data, type = "prob")

roc_analysis = roc(
  response  = test_data$readmission,
  predictor = final_pred_prob[, "Yes"]
)

plot(
  1 - roc_analysis$specificities,
  roc_analysis$sensitivities,
  type = "l", col = "steelblue", lwd = 2,
  ylab = "Sensitivity",
  xlab = "1 - Specificity",
  main = "Figure 5: ROC Curve — Final Model (Test Set)"
)
abline(a = 0, b = 1, lty = 2)
legend("bottomright",
       legend = paste("AUC =", round(auc(roc_analysis), 3)),
       col    = "steelblue", lwd = 2)

cat("Test Set AUC:", round(auc(roc_analysis), 3), "\n")
```

**Variable Importance — Final Model**

```{r varimp, message=FALSE}
plot(varImp(final_model), top = 10,
     main = "Figure 6: Final Model — Top 10 Most Important Features")

varImp(final_model)
```

**Description:** In the test set, the final Random Forest model achieved an accuracy of
[*X*]%, sensitivity of [*X*]%, specificity of [*X*]%, and Kappa of [*X*]. The AUC of
[*X*] in _Figure 5_ confirms [*good/moderate*] discriminative ability above the 0.5
baseline of random guessing. As shown in _Figure 6_, the most important predictors were
`WBC`, `Age`, and `sodium`, which is clinically meaningful — elevated white blood cell
count at admission signals acute systemic inflammation that can impair post-discharge
recovery, older age reflects cumulative cardiovascular vulnerability, and low serum sodium
(hyponatremia) is a known independent predictor of adverse outcomes in cardiac patients.
The low PPV ([*X*]%) reflects the underlying rarity of the outcome (9.4% readmission
rate); the model is better suited as a **screening tool** to flag high-risk patients for
targeted follow-up interventions than as a stand-alone diagnostic instrument.

---

#### **3. Reflection: Comparing Random Forest and GBM in a Clinical Setting**

In this analysis, we compared two ensemble tree-based methods — random forest and gradient
boosting — to predict 30-day AMI readmission. Briefly discuss one key conceptual difference
between these two methods and how that difference may affect their suitability for clinical
prediction.

**Answer:** Random forest and gradient boosting differ fundamentally in how trees are built
and combined. In random forest, trees are grown *in parallel* and independently on
bootstrapped samples of the training data, and predictions are averaged across all trees.
This parallel structure makes random forest robust to noise and less prone to overfitting,
since errors in individual trees tend to cancel out. In gradient boosting, trees are grown
*sequentially*, with each new tree specifically targeting the residual errors left by the
previous model. This allows GBM to reduce bias systematically and capture complex patterns
in the data, but also makes it more sensitive to hyperparameter choices and more
computationally demanding to tune. In a clinical prediction context with a relatively small
sample size — as in this AMI readmission dataset — random forest's greater stability may be
advantageous, whereas GBM's flexibility can yield higher performance when there is
sufficient data and careful tuning. The choice ultimately depends on whether the clinical
priority is robustness and consistency or maximizing discriminative performance.
